### 2015-02-27 src/4_OnlineLearning.tex
demo bias-variance tradeoff

\begin{align}
\mathrm{E}\Big[\big(y - \hat{f}(x)\big)^2\Big]
 & = \mathrm{Bias}\big[\hat{f}(x)\big]^2 + \mathrm{Var}\big[\hat{f}(x)\big] + \sigma^2
\end{align}
\noindent where:
\begin{align*}
 \mathrm{Bias}\big[\hat{f}(x)\big] &= \mathrm{E}\big[\hat{f}(x)\big] - f(x) \\
 \mathrm{Var}\big[\hat{f}(x)\big] &= \mathrm{E}\Big[ \big( \hat{f}(x) - \mathrm{E}[\hat{f}(x)] \big)^2 \Big] 
\end{align*}

@@

The optimal solution for $\mathbf{\mathbf{X}}$ is:
\begin{equation*}
\mathbf{\mathbf{X}}= \mathbf{A}^+ \mathbf{Y}
%(\mathbf{X}^\top\mathbf{X}
%\mathbf{X})^{-1}\mathbf{X}^\top \mathbf{Y}
\end{equation*}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\noindent where $\mathbf{A}^+$ is the Moore-Penrose pseudo-inverse
which can be written as follows:

@@

\begin{equation*}
\label{eq:problem} 
J(\mathbf{\mathbf{X}}) =  \| \mathbf{A}\mathbf{\mathbf{X}} - \mathbf{Y} \|_2^2  + \lambda
 \| \mathbf{\mathbf{X}}\| ^2
\end{equation*}

@@

@@

>>>>

\begin{align}
\mathrm{E}\Big[\big(y - \hat{f}(x)\big)^2\Big]
 & = \mathrm{Bias}(\hat{f}(x))^2 + \mathrm{Var}(\hat{f}(x)) + \sigma^2
\end{align}
\noindent where:
\begin{align*}
 \mathrm{Bias}(\hat{f}(x)) &= \mathrm{E}\big[\hat{f}(x)\big] - f(x) \\
 \mathrm{Var}(\hat{f}(x)) &= \mathrm{E}\Big[ \big( \hat{f}(x) - \mathrm{E}[\hat{f}(x)] \big)^2 \Big] 
\end{align*}

\begin{eqnarray*}
E[(y-\hat{f}(\mathbf{x}))^2] & =&
E[(y-f(\mathbf{x})+f(\mathbf{x})-\hat{f}(\mathbf{x}))^2] \\
&=& E[(B-f(\mathbf{x}))^2] +
E[(f(\mathbf{x})-\hat{f}(\mathbf{x}))^2] + \dots \\
& &
2\cancelto{0}{E[y-f(\mathbf{x})]}E[f(\mathbf{x})-\hat{f}(\mathbf{x})] \\
&=& \sigma^2 + MSE(\hat{f}(\mathbf{x}))
\end{eqnarray*}

\begin{eqnarray*}
    MSE(\hat{f}(\mathbf{x})) &=&
    E[f(\mathbf{x})-\hat{f}(\mathbf{x}))]^2 \\
    &=& E[f(\mathbf{x})-E[\hat{f}(\mathbf{x})] +
    E[\hat{f}(\mathbf{x})]-\hat{f}(\mathbf{x}))]^2 \\
    &=&  E[(f(\mathbf{x})-E[\hat{f}(\mathbf{x})])^2] +
      E[(\hat{f}(\mathbf{x})-E[\hat{f}(\mathbf{x})])^2] + \dots\\
    & & E[(f(\mathbf{x})-E[\hat{f}(\mathbf{x})])] 
    \cancelto{0}{E[(\hat{f}(\mathbf{x})-E[\hat{f}(\mathbf{x})])]} \\
    &=& Bias(\hat{f}(\mathbf{x}))^2 + Var(\hat{f}(\mathbf{x}))
\end{eqnarray*}

@@

The optimal solution for $\mathbf{\mathbf{X}}$ is:
\begin{equation}
\label{eq:MP}
\mathbf{\mathbf{X}}= \mathbf{A}^+ \mathbf{Y}
\end{equation}

\noindent where $\mathbf{A}^+$ is the Moore-Penrose pseudo-inverse
which can be written as follows:

@@

\begin{equation}
\label{eq:RRproblem} 
J(\mathbf{\mathbf{X}}) =  \| \mathbf{A}\mathbf{\mathbf{X}} - \mathbf{Y} \|_2^2  + \lambda
 \| \mathbf{\mathbf{X}}\| ^2
\end{equation}

@@

\subsection{Selection of Lambda}

The additional term $\lambda \|\mathbf{\mathbf{X}}\|_2^2$ in the optimisation
problem shown in equation~(\ref{eq:RRproblem}) has two effects on the solution:
shrinks the coefficients towards zero and improves the conditioning of the
problem.

@@

Figure~\ref{fig:shrinks} shows a visual example of the shrinking of the
coefficients:

\begin{figure}[h!]
\includegraphics[width=0.5\linewidth]{img/shrinks}
\caption{Shrink of regression coefficients}
\label{fig:shrinks}
\end{figure}

On the other hand, the effect of adding the term $\lambda \mathbb{I}$
to the matrix $\mathbf{A}^\top \mathbf{A}$
(equation~(\ref{eq:optsolRR})) improves its condition number since it
increases its diagonal values when $\lambda > 0 $.  The matrix
$\mathbf{A}^\top \mathbf{A}$ is symmetrical ($(\mathbf{A}^\top
\mathbf{A})^\top = \mathbf{A}^\top \mathbf{A}$) and therefore
diagonalizable.  If we know the eigenvalue decomposition of $\mathbf{A}
= \mathbf{U\Sigma V^\top}$, then:

\begin{eqnarray*}
\mathbf{A}^\top \mathbf{A}+\lambda \mathbb{I}&=&\mathbf{V\Sigma^2
V}^\top + \lambda \mathbf{V} \mathbf{V}^\top\\ &=&\mathbf{V}
(\mathbf{\Sigma}^2+\lambda\mathbb{I}) \mathbf{V}^\top \, ,
\end{eqnarray*}

\begin{equation*}
\mathbf{\Sigma}^2+\lambda\mathbb{I}=
\begin{bmatrix}
\sigma^2_1 + \lambda & \, & \, \\
\, & \sigma^2_2 +\lambda & \, \\
\, & \, & \ddots & \, \\
\, & \, & \, & \sigma^2_n +\lambda \, .
\end{bmatrix}
\end{equation*}

where $\sigma_1 \geq \sigma_2 \geq \dots \geq \sigma_n$.

Since the condition number of a matrix $\mathbf{A}$ is defined as:

\begin{equation*}
	\kappa = \|\mathbf{A}\| \|\mathbf{A}^{-1}\|
\end{equation*}

If matrix $\mathbf{A}$ is non-singular, its condition number can be
expressed in terms of its singular values. The effect of adding the
regularization term affects the condition number as follows:

\begin{eqnarray*}
\kappa_{ols} &=& \|\mathbf{A}\| \|\mathbf{A}^{-1}\|=\frac{\sigma_1}{\sigma_n} \\
\kappa_{ridge} &=& \|\mathbf{A}^\intercal \mathbf{A} + \lambda \mathbb{I}\| 
\|(\mathbf{A}^\intercal \mathbf{A} + \lambda \mathbb{I})^{-1}\|=\frac{\sigma_1+\lambda}{\sigma_n + \lambda} \,
\end{eqnarray*}

It is easy to see that the term $\lambda$ improves the condition number:

\begin{equation*}
        \frac{\sigma_1+\lambda}{\sigma_n + \lambda} <
        \frac{\sigma_1}{\sigma_n} \,  \qquad \forall \quad \lambda > 0
\end{equation*}

However, $\lambda$ cannot be too large. Tipically $\lambda$ is small and its
magnitude depends on the matrix $\mathbf{A}$.

For rank deficient matrices we know that $\text{det}(\mathbf{A A^\top})=0$, adding the
term $\lambda \mathbb{I}$ we have that $\text{det}(\mathbf{A A^\top}+\lambda \mathbb{I}) =
p(\lambda)$ where $p(\lambda)$ is a polynomial of degree $n$ ($\mathbf{A}$ is
$m \times n$). The zeros of $p(\lambda)$ are discretes, so it can be represented
as:

\[
p(\lambda) =
\lambda(\lambda-\lambda_1)^{n_1}(\lambda-\lambda_2)^{n_2}\dots(\lambda-\lambda_s)^{n_s}
\]

\noindent where $n_1 + n_2 + \dots + n_s = n$.

This means that $\lambda$ must be small in order to ensure that $p(\lambda)$
does not vanish.

%\subsection{The bias-variance tradeoff}
%
%To determine parameter $\lambda$ is crucial for ridge regression since it could
%reduce the expected prediction error by reducing variance (bias-variance trafeoff) considering a biased
%estimator. It is known that the prediction error is
%obtained as:
%
%\begin{equation}
% \label{eq:prederror}
%E[(B-\hat{f}(\mathbf{X}))^2] = \sigma^2 + Bias(\hat{f}(\mathbf{X}))^2 + Var(\hat{f}(\mathbf{X}))
%\end{equation}
%
%\noindent where $\hat{f}(\mathbf{X})=\mathbf{AX}(\lambda)$ and
%$f(\mathbf{X})=\mathbf{AX}$.
%
%
%
%The bias of OLS is:
%
%\begin{eqnarray*}
%Bias(\hat{\mathbf{X}}) &=& E[\hat{\mathbf{X}}] - \mathbf{X} \\
%&=& E[ (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{B}] - \mathbf{X} \\
%&=& E[ (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top (\mathbf{AX})] - \mathbf{X}  \\
%&=& \mathbf{X}  - \mathbf{X}  \\
%&=&  0
%\end{eqnarray*}
%
%
%The bias of ridge regression when $\mathbf{A A^\top}$ is non-singular
%can be obtained expressing ridge regression solution
%$\mathbf{\lambda}$ in terms of OLS solution $\hat{\mathbf{X}}$:
%
%\begin{eqnarray*}
%\mathbf{X}(\lambda) &=&( \mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I})^{-1}\mathbf{B} \\
%&=& (\mathbb{I} + \lambda (\mathbf{A}^\top \mathbf{A})^{-1})^{-1} (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{B} \\
%&=&  (\mathbb{I} + \lambda (\mathbf{A}^\top \mathbf{A})^{-1})^{-1}  \hat{\mathbf{X}} \\
%&=& \mathbf{W} \hat{\mathbf{X}} 
%\end{eqnarray*}
%
%\noindent where $\mathbf{W}  = (\mathbb{I} + \lambda (\mathbf{A}^\top
%\mathbf{A})^{-1})^{-1}  $ it is defined for simplicity. Ridge
%regression bias is then obtained as:
%
%\begin{eqnarray*}
%Bias(\mathbf{X}(\lambda)) &=& E[\mathbf{X}(\lambda)] - \mathbf{X} \\
%&=& E[\mathbf{W}\hat{\mathbf{X}}] - \mathbf{X} \\
%&=&  \mathbf{W} \mathbf{X} - \mathbf{X} \neq 0 
%\end{eqnarray*}
%
%
%
%The variance of OLS is:
%
%\begin{equation*}
%Var(\hat{\mathbf{X}}) = \sigma^2 (\mathbf{A}^\top \mathbf{A} )^{-1}
%\end{equation*}
%
%\noindent and the variance of ridge regression is:
%
%\begin{eqnarray*}
%Var(\mathbf{X}(\lambda)) &=& Var(\mathbf{W}\hat{\mathbf{X}}) \\ 
%&=& \mathbf{W}Var(\hat{\mathbf{X}})\mathbf{W}^\top \\
%&=& \sigma^2 \mathbf{W}(\mathbf{A}^\top \mathbf{A} )^{-1}\mathbf{W}^\top
%\end{eqnarray*}
%
%
%\begin{figure}[h!]
%\includegraphics[width=0.8\linewidth]{plots/biasvariance}
%\caption{Bias-variance tradeoff}
%\label{fig:biasvariance}
%\end{figure}
%
%The figure~\ref{fig:biasvariance} shows the bias-variance tradeoff
%given by equation~(\ref{eq:prederror}). Ridge regression shows an
%increasing bias and a decreasing variance. Despite OLS has zero bias,
%its variance is greater than ridge for small values of $\lambda$.
%It can be shown that, in terms of prediction error, ridge (black line)
%is lower than OLS (dotted line)~\cite{hoerl+kennard1970}.
%
%
%
